\section*{Contribution}

AutoSage comprises five integrated layers.

\textbf{Metrics Collection.}
The system connects via \texttt{kubeconfig} and collects telemetry every
300\,s through a sequential fallback chain: \texttt{kubectl~top},
\texttt{metrics.k8s.io} API, and Kubelet summary endpoint. No Prometheus
deployment is required.

\textbf{Predictive Monitoring \& Uncertainty Quantification.}
CPU is forecast with a linear-trend plus seasonal composite; memory with
exponential smoothing ($\alpha\!=\!0.3$). Each forecast is wrapped in a
bootstrap prediction interval ($B\!=\!50$ resamples) whose total variance
grows with the horizon:
$\sigma^2(h) = \sigma^2_\mathrm{ale}\!\cdot\!\sqrt{h} + \sigma^2_\mathrm{epi}\!\cdot\!h$,
yielding 95\% intervals that naturally widen. Exceedance probabilities
$\Pr[Y\!>\!\tau]$ are computed for $\tau\!\in\!\{50,70,80,90\}\%$.
Anomaly detection uses an Isolation Forest calibrated via Platt scaling
($P = \sigma(As{+}B)$, $A\!=\!{-}5$, $B\!=\!{-}0.5$) into a four-level
severity distribution.

\textbf{LLM Autoscaling Advisor.}
A structured prompt (metrics, forecasts, statefulness, HPA/VPA status) is
sent to an LLM that returns a JSON recommendation (action, target,
confidence, reasoning). A two-tier fallback chain uses a self-hosted model
(Qwen, 240\,s timeout, $\theta\!=\!0.1$) with Groq (Llama~3.1-8B, 15\,s) as
backup. The LLM selects HPA for stateless and VPA for stateful workloads; an
enforcement layer corrects contradictions automatically.

\textbf{MCDA Cross-Validation.}
Every LLM recommendation is validated by a TOPSIS module that generates 5--8
candidate replica counts and scores each on five criteria (cost, performance,
stability, forecast alignment, response time) under four selectable weight
profiles. Agreement is classified as \emph{full}, \emph{partial}, or
\emph{divergent}; if the MCDA score exceeds the LLM's by $\delta\!>\!0.15$,
the system overrides the LLM choice.

\textbf{Dashboard \& Actuation.}
A Flask web dashboard with WebSocket support visualises prediction intervals,
exceedance bars, uncertainty decomposition (aleatoric vs.\ epistemic), anomaly
severity, and MCDA validation. Approved recommendations are applied via
\texttt{kubectl~patch}.
