% abstract
Modern cloud-native applications deployed on Kubernetes rely on reactive autoscalers
such as the Horizontal Pod Autoscaler (HPA) and Vertical Pod Autoscaler (VPA) to
handle dynamic workloads. These mechanisms act only after performance thresholds are
crossed, leading to provisioning lag, SLA violations, and resource over-commitment.
Furthermore, no existing autoscaler provides an auditable, human-readable explanation
for each scaling decision.

This thesis presents \textbf{Ai4k8s}, an agentic AI platform for intelligent and
explainable Kubernetes operations. Ai4k8s replaces reactive threshold logic with
a three-layer decision pipeline:
(1) an enforcement layer that applies hard safety rules, (2) a local Large Language
Model (LLM) that reads live cluster metrics and produces a plain-English scaling
recommendation, and (3) a Multi-Criteria Decision Analysis (MCDA) optimizer based
on TOPSIS that weighs cost, performance, stability, and forecast alignment to confirm
or override the LLM recommendation.

The platform is deployed entirely on-premises using \textit{qwen3:0.6b} (522~MB)
via Ollama, achieving 1.7~s inference at idle with no cloud API dependency. Under
CPU-saturating workloads a graceful cascade falls back to a cloud LLM (Groq) in
under 2~s, ensuring continuous operation.

We evaluate Ai4k8s against native HPA and VPA on a three-service muBench
microservice chain under realistic HTTP burst load. Ai4k8s reduces compute cost
by \textbf{52\%} compared to HPA at equivalent load (cost proxy 0.184 vs.\ 0.380),
while producing a full decision trace—enforcement verdict, LLM reasoning, and MCDA
score breakdown—for every scaling action. VPA achieves the lowest latency (p95
= 1.09~s, 0\% SLA violations) by right-sizing pod CPU limits, while Ai4k8s
occupies a distinct position: explainability and cost efficiency at the expense
of raw latency.

The work demonstrates that a sub-gigabyte local LLM, combined with a lightweight
MCDA optimizer, can drive safe, cost-effective autoscaling decisions on a single-node
Kubernetes cluster with commodity hardware, paving the way for fully on-premises
agentic AIOps.